---
title: 'Santa Clara River Phenology'
author: "{Kristina Glass}"
date: "2023-12-15"
output:
    html_document:
      print_df: paged
      toc: yes
      toc_depth: 4
      toc_float: yes
---

# Introduction

*\*Credit: This project is based on materials developed by Chris Kibler*

## What is Phenology

Phenology is the cyclical and seasonal lifecycle of natural phenomenon. For plants, importent phenological events include their growth of leaves, flowering, and death of leaves (or senescence). Some common phenological strategies of plants include the loss of leaves in winter and growth of leaves in spring (winter deciduous), the loss of leaves in summer (drought deciduous), and the maintenance of leaves year-round (evergreen). Due to climate change, plant phenology is shifting and it is important to understand what these shifts include and where these shifts are happening.

## Area of Interest

For this study, we're interested in exploring plant phenology near the Santa Clara River which flows from Santa Clarita, CA and Ventura, CA. It is the last major coastal system within Southern California that has been relatively unchanged by development and supports over 17 species. The Santa Clara River is also supportive of one of the most valuable agricultural in the world[1](http://rivrlab.msi.ucsb.edu/santa-clara-river).

# Purpose of Study

The purpose of this study is to investigate plant phenology of riparian forests, grasslands, and chaparral shrub lands along the Santa Clara River. Riparian forests grow along the river and are dominated by winter deciduous cottonwood and willow trees, grasslands grow in open spaces and are predominate dominated by deciduous grasses, and chaparral shrub lands grow in more arid areas and are dominated by evergreen shrubs.

We will explore the vegetation productivity (NDVI), calculate the NDVI throughout the year, summarize the NDVI values within vegetation communities, and visualize the changes in NDVI within vegetation communities.

# Data

**Landsat Operational Land Imager (OLI sensor)** which contains 8 pre-processed scenes. The attributes of the scenes are as follows:

1.  Level 2 surface reflectance products

2.  erroneous values were set to NA

3.  Scale factor set to 100

4.  Bands are 2-7

5.  Dates of the scenes are in file name

and

**Study sites** which are polygons that represent the sites of interest. The attributes are as follows:

1.  "study_site" is a character string with plant type

The data are available in a folder labeled "data" within the repository

# Workflow

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, include = TRUE)
```

```{r include=TRUE, message=FALSE}
library(terra)
library(sf)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(here)
library(tmap)
library(cowplot)

rm(list = ls())
here::i_am("assignments/santa_clara_river_phenology/river_phenology.Rmd")
```

## Function to Compute Single NDVI Scene

This function computes the Normalized Difference Vegetation Index (NDVI). NDVI is a computation of the differences in reflectance (the ratio of incoming light reflected by a leaf) in NIR (near infrared) and red bands that are then normalized by sum

```{r ndvi-function}
# create function to compute single NDVI scene
ndvi_fun <- function(nir, red){
  
  (nir - red)/(nir + red)
  
}

```

```{r, warning=FALSE, messages=FALSE}
# read in first landsat scene from June 12, 2018
landsat_20180612 <- rast(here("assignments", "santa_clara_river_phenology","data", "landsat_20180612.tif"))
#view dataset
#landsat_20180612
#update layer names
names(landsat_20180612) <- c("blue", "green", "red", "NIR", "SWI1", "SWI2")
#confirm layer names were updated
#landsat_20180612
```

The NDVI function created above can be applied with the `lapp()` function, which applies a function to each cell using layers as arguments. To do so, the layers need to be passed into the function. For this data, NIR is the 4th layer and the red band is the 3rd. Bands or layers need to be applied in the order in which they were defined in the function.

```{r}
#compute NDVI for June 12th, 2018 with lapp()
ndvi_20180612<-lapp(landsat_20180612[[c(4,3)]], fun = ndvi_fun)
#view dataset
ndvi_20180612
#exploratory output plot of dataset
plot(ndvi_20180612)
```

## Function to Compute all NDVI Scenes

To make a more fluid process to compute NDVI, I created a function that computes all NDVI scenes simultaneously

```{r, warning=FALSE, messages=FALSE}
rm(list = ls())
here::i_am("assignments/santa_clara_river_phenology/river_phenology.Rmd")

ndvi_fun = function(nir, red){
  (nir - red) / (nir + red)
}
```

```{r eval=FALSE}
# create a basic outline of the function
# note: this code is not meant to run! we're just outlining the function we want to create
create_ndvi_layer <- function(){
  landsat <- rast(file) # read in scene
  names(landsat) <- c("blue", "green", "red", "NIR", "SWIR1", "SWIR2") # rename layers
  ndvi <- lapp(landsat[[c(4, 3)]], fun = ndvi_fun) # compute NDVI
}

```

```{r, warning=FALSE, messages=FALSE}

#calling for a list if files from the data folder stores all data to one variable
# the "pattern" option returns names that end in .tif (the file extension for the landsat scences)
# the "full.names" option returns the full file path for each scene

files <- list.files(here("assignments", "santa_clara_river_phenology","data"), pattern = "*.tif", full.names = TRUE)
```

```{r, warning=FALSE, messages=FALSE}
#update function to work with list of file names
# now we're passing our function a number that will correspond to the index in the list of file names
create_ndvi_layer <- function(i){
  landsat <- rast(files[i])
  names(landsat) <- c("blue", "green", "red", "NIR", "SWIR1", "SWIR2")
  ndvi <- lapp(landsat[[c(4, 3)]], fun = ndvi_fun)
}

#testfunction by asking it to read in the first file 
test <- create_ndvi_layer(1)
```

```{r, warning=FALSE, messages=FALSE}
#run function on all scenes
# create NDVI layer for each scene (day) and stack into a single rasterstack
all_ndvi <- c(create_ndvi_layer(1),
              create_ndvi_layer(2),
              create_ndvi_layer(3),
              create_ndvi_layer(4),
              create_ndvi_layer(5),
              create_ndvi_layer(6),
              create_ndvi_layer(7),
              create_ndvi_layer(8))

# update layer names to match date
names(all_ndvi) <- c("2018-06-12",
                     "2018-08-15",
                     "2018-10-18",
                     "2018-11-03",
                     "2019-01-22",
                     "2019-02-23",
                     "2019-04-12",
                     "2019-07-01")
```

## Compare NDVI Across Vegetation Communities

Since the NDVI for each scene was computed, I can now compare changes in NDVI values across the vegetation communities within the region. As a reminder, those communities are: riparion forests, grasslands, chaparral shrub lands

Now that we have computed NDVI for each of our scenes (days) we want to compare changes in NDVI values across different vegetation communities.

```{r include=TRUE}
#read in shapefile of the study sites
sites <- st_read(here("assignments", "santa_clara_river_phenology","data", "study_sites.shp"))

# plot study sites on a single NDVI layer
tm_shape(all_ndvi[[1]]) +
  tm_raster() +
  tm_shape(sites) +
  tm_polygons()
```

Now, I will find the

Here we find the average NDVI within each study site. The output of extract is a data frame with rows that match the study site dataset, so we bind the results to the original dataset.

```{r include=TRUE}
#extract NDVI at Study Sites
#find average NDVI within each study sites
sites_ndvi <- terra::extract(all_ndvi, sites, fun = "mean")
#output of sites_ndvi will be a data frame that has rows that match the study sites data set

#bind both datasets by column
sites_annotated <- cbind(sites, sites_ndvi)

#clean dataset
sites_clean <- sites_annotated %>%
  st_drop_geometry() %>% #remove unecessary geometry
  select(-ID) %>% #remove ID column
  pivot_longer(!study_site) %>% #increases number or rows and decreases number of columns
  rename("NDVI" = value) %>% #rename column
  mutate("year" = str_sub(name, 2, 5), #extracts and replaces elements at a single position. finds emelent position within column and places it into new columns labled by year, month, day
         "month" = str_sub(name, 7, 8),
         "day" = str_sub(name, -2, -1)) %>%
  unite("date", 4:6, sep = "-") %>% #mutated columns above are brought into a single date column, separated by "-"
  mutate("date" = lubridate::as_date(date)) %>% #turn date column into a datetime object
  select(-name) #remove names column
```

## Plot Results

```{r include=TRUE}
ggplot(sites_clean,
       aes(x = date, y = NDVI,
           group = study_site, col = study_site)) + #plot data colored by study_site
  scale_color_manual(values = c("#EAAC8B", "seagreen", "darkolivegreen", "#9EA93F","goldenrod")) +
  geom_line() + #plot NDVI as line
  theme_dark() + #dark background so its easier to see colors in line
  labs(x = "", y = "Normalized Difference Vegetation Index (NDVI)", col = "Vegetation type",
       title = "Seasonal cycles of vegetation productivity")
```

NDVI values will range from -1.0 - 1.0, where negative values indicate clouds and water, positive values near zero indicate bare soil, values from 0.1-0.5 indicating sparse vegetation and 0.6-1.0 indicating dense green vegetation [2](https://ipad.fas.usda.gov/cropexplorer/Definitions/spotveg.htm). With this knowledge, we can begin to interpret the results.

### Chaparral

NDVI primarily remained above 0.6, save for a dip falling to 0.55 at the lowest between July and October. Since these vegetation communities are dominated by evergreen species, this is an expected result since they stay green all year long. The small dip in NDVI could be attributed to high temperatures seen during the summer or a small drought period.

### Riparian Forests

Forest1 and Forest 2 maintain similar cycles of productivity, experiencing drops in NDVI under 0.5-0.4 from January to April, and rises in productivity during summer months. Since the communities within Riparian forests are winter deciduous, they will lose their leaves during the fall and winter and regrow them through the spring in summer, explaining the rise and fall cycles of its NDVI.

For Forest3 maintains similar productivity cycles as those listed above, but has much longer periods of maintaining sparse vegetation, save for a spike above 0.6 after April 2019.

### Grassland

This vegetation experiences the most extreme shifts in NDVI, with values under 0.3 from July 2018 to just after October 2018 where it experiences a sharp spike, followed by a sudden decline after April 2019. Since these grasses are deciduous, they lose their leaves in the winter and gain them back during the Spring.

# Conclusion

From the data available, we can see that in 2019, NDVI values as of July were above NDVI values for the following year. This could be due to longer periods of rain[3](https://www.latimes.com/california/story/2019-12-31/southern-california-precipitation-above-normal-year-end), longer periods of warm weather, or other climate related factors, which would be worth investigating for future study. While none of the study sites experienced extreme deviations expected NDVI from 2018-2019, the shift in their seasonal productivity cycle is present as indicated by the plot above.

# References

Duginski, P. (2020, January 1). *At Year's End, Southern California's precipitation remains above normal*. Los Angeles Times. <https://www.latimes.com/california/story/2019-12-31/southern-california-precipitation-above-normal-year-end>

*Normalized difference vegetation index (NDVI)*. International Production Assessment Division (IPAD) - Home Page. (n.d.). <https://ipad.fas.usda.gov/cropexplorer/Definitions/spotveg.htm>

*Santa Clara River Research and Restoration*. Rivrlab.msi.ucsb.edu. (n.d.). <http://rivrlab.msi.ucsb.edu/santa-clara-river>
